On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-17 manuscripts/PNAS/review/point_by_point_response.txt

-----------------
Since the result is an extension of the noise-free result of (HS15) to a noisy case, the authors should provide better justification for the significance of contributions. The main result seems to be based on an approach similar to (HS15), and there is little to no discussion of the computability and tightness of the sufficient condition for stability in the main results. Thus, the submission does not seem to be greatly more valuable than (HS15) and also not conclusive regarding the central problems of sparse linear coding and dictionary learning. That being said, the paper (with the modifications below) would certainly merit publication in very solid journal like IEEE Transactions on Information Theory, where it would be a median paper, in my opinion.
-----------------

RESPONSE:

We agree with the reviewer that a straightforward extension of the approach and noise-free result of (HS15) to the noisy case — while still a significant advancement in applied mathematics with actual practical implications — would not necessarily merit publication in PNAS on its own. The reviewer, however, seems to have overlooked the fact that our main result (Thm. 1) generalizes the noise-free result of (HS15) far beyond a simple extension to the noisy case in ways which provide valuable insight into the dictionary learning model and that a significant deviation from the approach taken in (HS15) was necessary to prove these new facts. Moreover, our solution (Cor. 2) to the optimization formulation of the dictionary learning problem (Prob. 2) — that of most interest to optimization theorists and practitioners of their methods — is (to our knowledge) the first of its kind in both the noise-free and noisy domains. We discuss all of these points in more detail in our response to the reviewer’s later "detailed comments".

We have included in our updated manuscript a discussion of the computability of the sufficient conditions (they are, indeed, computable), as per the reviewer’s request. The main constants in our work -- C_1, C_2 -- are computable, elementarily, in terms of two positive quantities that are natural in our context: (a) the restricted “Lower Bound" L_k(M) of a matrix M, which is a measure of how distinguishable two k-sparse x’s are by their images under M, and (b) the value in Def. 4 of \xi(V_1,…,V_l) given subspaces V_i spanned by columns of A, which bounds the distance between intersections of such subspaces and each subspace individually. The first quantity (a) is, in some form or another, related to standard objects in the field and may be computed as the smallest singular value of a set of specific sub-matrices; in particular, for fixed sparsity constraint k, the lower bound L_k(A) is efficiently computable (polynomial time). Because deciding L_k(A) > 0 is already NP-hard (TP14), it seems likely that without fixing a sparsity level there is no efficient computation in general for (a). The calculation of (b) involves determining "canonical angles" between subspaces, which is usually reduced again to a numerical (efficient) singular value decomposition given the fixed subspaces.  However, upon close inspection, C_2 nonetheless requires an exponential number of instances of computing \xi. Thus, while our constants are effective (an algorithm exists to compute them), they are not necessarily polynomial-time computable.

@@

To be clear, the new perspective we take on the problem yields the following powerful conclusions beyond those of a straightforward extension of (HS15) to the noisy case:

1) An extension to the case where the number of dictionary elements is unknown: The results of (HS15) only apply to the case where the matrix B has the same number of columns as A. We forgo this assumption and show that B must have at least as many columns as A and contains (up to noise) a subset of the columns of A. The size of this subset depends on a simple relation between the number of columns of B and the regularity of the support-set hypergraph.

2) A significant reduction in sample complexity: To identify the n x m generating dictionary A, (HS15) require that data be sampled from every k-dimensional subspace spanned by the m columns of A (that is, {m \choose k} subspaces in total). We show that the data need only be sampled from m subspaces in total (e.g. those supported by consecutive intervals of length k in some cyclic order on [m]) and in some cases as few as 2\sqrt{m} (e.g. when m = k^2, take the supports to be the rows and columns of [m] arranged in a square grid). Moreover, if the size of the support set of reconstructing codes is known to be polynomial in m and k, then the pigeonholing argument in the proof of Thm. 1 requires only a polynomial number of samples distributed over a polynomial number of supports; thus, N is polynomial in m, k. This point was only hinted at in the Discussion section of our original submission, but we have included it in the updated manuscript to make clear the power of our approach over that taken in (HS15). We are aware of only one work treating combinatorial hypergraph designs in the setting of dictionary learning, "An incidence geometry approach to dictionary learning" (arxiv, non-published paper)

3) No spark condition requirement for the generating matrix A: One of the mathematically significant contributions made by (HS15) was to forgo the constraint that the recovery matrix B also satisfy the spark condition, in contrast to all previously known uniqueness results which (either explicitly or implicitly) made this assumption.  Our proof is powerful enough to show that, in fact, even the matrix A need not satisfy the spark condition to be identifiable! Rather, it need only be injective on the union of subspaces with supports that form sets in a hypergraph satisfying the singleton intersection property (SIP). (For example, consider the matrix A = [e_1, …, e_5, v] where v = e_1 + e_3 + e_5. and take H to be the set of all consecutive pairs of [m] arranged in cyclic order. Then A satisfies the assumptions of Thm. 1 without satisfying the spark condition.) We had omitted this point in our original submission of the manuscript to keep things simple, but have decided to include this fact in our resubmission. This also required us to redefine the restricted matrix lower bound L_k to be in terms of a hypergraph H (L_H in the revised manuscript), which is an interesting object for further study in its own right.

@@

RESPONSE: Yes, thanks for this question. The condition on \ell is perhaps best understood as a condition on \bar m not being too large (given \ell) in order to guarantee the recovery of at least one dictionary element. We have updated the manuscript to reflect this, and the condition now reads “provided \bar m < \ell m / (\ell - 1)” where previously it read “provided p = \bar m − \ell( \bar m − m ) is positive”.

-----------------
-- The paper could also compare the sample complexity results on N with those given in the noise-free case. For example, Table I in reference (18) gives a comparison of sample complexity requirements for different conditions in the noise-free case. Where would the results of this paper stand in that table?

-- Otherwise, all the math seemed correct.
-----------------

RESPONSE: In brief, since all of the theorems and corollaries in (HS15) are corollaries of our theorems, all sample complexities improve by our work. The main point of difference between our statements about sample complexity and those of (HS15) is the assumed constraint on the underlying support set of sparse codes x_i. Our theory of combinatorial designs (the ‘singleton intersection property’) requires only that there be a sufficient number of x_i drawn from each support in a hypergraph satisfying the SIP, whereas the theory in (HS15) requires that a sufficient number of x_i be drawn from *every* support of size k in [m]. Below, we make an explicit comparison with Table I in (HS15) row by row, assuming for our results that sufficient data have been sampled from all supports in a hypergraph H satisfying the SIP:

@@

ROW I. Our result here is |H|(k-1){m \choose k} + 1. We improve on the result in (HS15) by an exponential factor since our theory does not require that data be sampled with every possible support of size k in [m]; as noted in our manuscript, for every k < m there exists a hypergraph H with |H| = m that satisfies the SIP.

ROW 2. Our result here is again |H|(k-1){m \choose k} + 1 with *certainty*, not with *almost certainty* (i.e. probability 1) as in (HS15). Here the authors in (HS15) have applied probabilistic arguments to achieve an almost certain result with a sample complexity on the same order as that for which we have achieved a certain result by means of our theory of combinatorial designs.

ROW 3. We cannot make a direct comparison here because we have not calculated the probability that a random set of supports satisfy the SIP (see our response to a similar question by the second reviewer). We think this is an interesting problem for the community to solve; regardless, the result of (HS15) here is still implied by our more general theory.

ROW 4. Our result here is |H|(k-1){m \choose k} + 1 with probability 1. Here is the only case where the sample complexity stated in (HS15) is technically better than ours, but it differs in flavor: their x_i are assumed to be distributed as (k+1) samples per support for every support of size k in [m], whereas ours supposes that (k-1){m \choose k} codes x_i are distributed over each support in some hypergraph H satisfying the SIP. Regardless, the (technically) better result of (HS15) is still implied by our more general theory in the case \varepsilon = 0 with the addition of their probabilistic argument.

@@

ROW 5: Again, we cannot make a direct comparison here for the reason stated in ROW 3.

To sum up, while we appreciate this question, we feel that since our results improve in every case except for one noise-free technicality (ROW 4), and since the results of (HS15) are all entailed anyway by our more general theory, an explicit comparison such as that which we provided above would not be the best use of our limited space in the manuscript; although we are willing to add a discussion about this if necessary. Many thanks for your careful review of the mathematics. We have tried and continue to try to be as clear, concise, and correct as we possibly can to elevate this work into the top echelon of applied mathematical theory papers.

@@

When is it possible to determine (numerically) if enough data has been gathered to obtain an accurate estimate of the underlying sparsity dictionary?
-----------------

RESPONSE: A statement of this very general nature requires that we calculate the probability that all of the conditions of Thm. 1 are satisfied by random data. We have provided an "almost certain" guarantee of this flavor in Cor 3, wherein the sparse vector supports are constrained to form a hypergraph satisfying the SIP. It would be useful and interesting to calculate furthermore the probability that random supports form a hypergraph satisfying the SIP. We have opted not to include these calculations due in part to space constraints, but also because we believe that what sets our work apart from the vast majority of results in the field is their deterministic nature, e.g. they do not depend on any kind of assumption about the random distribution from which the sparse coefficients or dictionary entries are drawn.

-----------------
Are the conditions and constants/metrics provided in this paper feasible to compute?
-----------------

RESPONSE: We have determined explicit forms for the effectively computable constants in Theorem 1, but they are not necessarily feasible to compute. We have included discussion of this in the revised manuscript.

### 2017-08-18 manuscripts/PNAS/chazthm_pnas.tex
sml things

\begin{theorem}\label{DeterministicUniquenessTheorem}
Fix an $n \times m$ matrix $\mathbf{A}$ with $L_{2\mathcal{H}}(\mathbf{A}) > 0$ for an $r$-regular $\mathcal{H} \subseteq {[m] \choose k}$ with the SIP. If \mbox{$\{\mathbf{x}_1, \ldots, \mathbf{x}_N\} \subset \mathbb{R}^m$} contains, for each $S \in \mathcal{H}$, more than $(k-1){m \choose k}$ $k$-sparse vectors in general linear position supported in $S$, then there is $C_1 > 0$ with the following holding for all\footnote{The condition $\varepsilon < L_2(\mathbf{A}) /C_1$ is necessary; otherwise,  with \mbox{$\mathbf{A}$ = $\mathbf{I}$} and $\mathbf{x}_i = \mathbf{e}_i$, there is a matrix $\mathbf{B}$ and $1$-sparse $\mathbf{\bar{x}}_i$ with $\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar{x}}_i \|_2 \leq \varepsilon$, nonetheless, violating \eqref{Cstable}.} $\varepsilon < L_{2}(\mathbf{A}) / C_1$:

@@

The total number of $(k-1)$-sparse vectors $\mathbf{\bar x}_i$ can thus not exceed $|\mathcal{H}|(k-1){ \bar m \choose k-1}$. By \eqref{eqn} no more than $|\mathcal{H}|k(k-1){ \bar m \choose k-1}$ vectors $\mathbf{\bar x}_i$ are not $k$-sparse. Since for every $S \in \mathcal{H}$ there are over $(k-1)\left[ {\bar m \choose k} + |\mathcal{H}|k{ \bar m \choose k-1} \right]$ vectors $\mathbf{x}_i$ supported there, it follows that more than $(k-1){\bar m \choose k}$ of them have corresponding $\mathbf{\bar x}_i$ that are $k$-sparse. The result now follows from Thm.~\ref{DeterministicUniquenessCorollary}.
\end{proof}

@@

The goal of this work was to explain the emergence of characteristic representations from sparse linear coding models fit to natural data despite the varied assumptions underlying the many algorithms in current use. To this end, we have taken an important step toward unifying the hundreds (if not thousands) of publications on the topic by demonstrating very general, deterministic conditions under which the identification of parameters in this parsimonious model is not only possible but also robust to the inevitable uncertainty permeating measurement and model choice.

@@

Beyond an original extension of existing noiseless guarantees \cite{Hillar15} to the noisy regime, a major innovation in our work is a theory of combinatorial designs for support sets key to the identification of the dictionary. We incorporate this idea into a new fundamental lemma in matrix theory (Lem.~\ref{MainLemma}) which draws on the definition of a new matrix lower bound induced by a hypergraph satisfying a ``singleton intersection property''. Insights enabled by our combinatorial approach include: 1) a subset of dictionary elements is recoverable even if dictionary size is overestimated, 2) data require only a polynomial number of distinct sparse supports, and 3) the spark condition is not a necessary property of recoverable dictionaries.

>>>>

\begin{theorem}\label{DeterministicUniquenessTheorem}
Fix an $n \times m$ matrix $\mathbf{A}$ with $L_{2\mathcal{H}}(\mathbf{A}) > 0$ for an $r$-regular $\mathcal{H} \subseteq {[m] \choose k}$ with the SIP. If \mbox{$\{\mathbf{x}_1, \ldots, \mathbf{x}_N\} \subset \mathbb{R}^m$} contains, for each $S \in \mathcal{H}$, more than $(k-1){m \choose k}$ $k$-sparse vectors in general linear position supported in $S$, then there is $C_1 > 0$ with the following holding for all\footnote{The condition $\varepsilon < L_2(\mathbf{A}) /C_1$ is necessary; otherwise,  with \mbox{$\mathbf{A}$ = $\mathbf{I}$} and $\mathbf{x}_i = \mathbf{e}_i$, there is a matrix $\mathbf{B}$ and $1$-sparse $\mathbf{\bar{x}}_i$ with $\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar{x}}_i \|_2 \leq \varepsilon$ that nonetheless violates \eqref{Cstable}.} $\varepsilon < L_{2}(\mathbf{A}) / C_1$:

@@

The total number of $(k-1)$-sparse vectors $\mathbf{\bar x}_i$ can thus not exceed $|\mathcal{H}|(k-1){ \bar m \choose k-1}$. By \eqref{eqn}, no more than $|\mathcal{H}|k(k-1){ \bar m \choose k-1}$ vectors $\mathbf{\bar x}_i$ are not $k$-sparse. Since for every $S \in \mathcal{H}$ there are over $(k-1)\left[ {\bar m \choose k} + |\mathcal{H}|k{ \bar m \choose k-1} \right]$ vectors $\mathbf{x}_i$ supported there, it follows that more than $(k-1){\bar m \choose k}$ of them have corresponding $\mathbf{\bar x}_i$ that are $k$-sparse. The result now follows from Thm.~\ref{DeterministicUniquenessCorollary}.
\end{proof}

@@

The goal of this work was to explain the emergence of characteristic representations from sparse linear coding models fit to natural data, despite the varied assumptions underlying the many algorithms in current use. To this end, we have taken an important step toward unifying the hundreds (if not thousands) of publications on the topic by demonstrating very general, deterministic conditions under which the identification of parameters in this parsimonious model is not only possible but also robust to the inevitable uncertainty permeating measurement and model choice.

@@

Beyond an original extension of existing noiseless guarantees \cite{Hillar15} to the noisy regime, a major innovation in our work is a theory of combinatorial designs for support sets key to the identification of the dictionary. We incorporate this idea into a new fundamental lemma in matrix theory (Lem.~\ref{MainLemma}) that draws upon the definition of a new matrix lower bound induced by a hypergraph satisfying a ``singleton intersection property''. Insights enabled by our combinatorial approach include: 1) a subset of dictionary elements is recoverable even if dictionary size is overestimated, 2) data require only a polynomial number of distinct sparse supports, and 3) the spark condition is not a necessary property of recoverable dictionaries.
